A Solution to Static vs. Dynamic Linking

### 2024-03-09 mss.tex
Post new style.
Post all.

\newcommand{\authorname}{Ted Blackman \patp{rovnys-ricfer}; Philip C. Monk \patp{wicdev-wisryt}}
\newcommand{\authorpatp}{}
\newcommand{\affiliation}{Urbit Foundation; Tlon Corporation}

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              Manuscript submitted for review.\\
              % {\it Urbit Systems Technical Journal} I:1 (2024):  1–3. \\
              ~ \\
              Address author correspondence to \patp{rovnys-ricfer}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{A Solution to Static vs. Dynamic Linking}
\fancyfoot[LE,RO]{\thepage}

@@

\begin{abstract}
Computing systems that utilize library code must either link to a self-provided version (static linking) or a system-supplied version (dynamic linking).  This can lead to memory duplication or dependency hell.  The Urbit \lstinline[style=inlinecode]{++ford} build system elegantly solves for the linking problem by promoting structural sharing of objects (nouns) in memory and by utilizing a referentially transparent build cache.
\end{abstract}

@@

\noindent
\citeauthor{Locklin2022}'s picturesque exposition highlights the “dependency hell” or “DLL hell” that mires modern software development.  (Cf. \citet{Grimes2003}.)  From Urbit's perspective as a solid-state computer, another problem with dynamic linking is that it is not deterministic.  Dynamic linking was something of a pact with the devil, permitting efficient memory usage at the price of legibility.

@@

As another example, the Hoon source file with text \lstinline[style=inlinecode]{'[3 0x4]'} compiles to \lstinline[style=inlinecode]{[[%cell [%atom %ud ~] [%atom %ux ~]] [3 0x4]]}.  It's a vase of a \lstinline[style=inlinecode]{%cell} (pair) of unsigned decimal number (\lstinline[style=inlinecode]{%ud}) and unsigned hexadecimal number (\lstinline[style=inlinecode]{%ux}).  If \lstinline[style=inlinecode]{/foo/hoon} is \lstinline[style=inlinecode]{'3'}, \lstinline[style=inlinecode]{/bar/hoon} is \lstinline[style=inlinecode]{'0x4'}, then importing both of those files changes the build subject to (a vase of) \lstinline[style=inlinecode]{[foo=3 bar=0x4 <stdlib>]}.

@@

In Urbit, everything is a “noun” (a binary tree with arbitrarily-sized integers at leaves).  If you “copy” a noun \lstinline[style=inlinecode]{foo}, like \lstinline[style=inlinecode]{[foo foo]}, the runtime just copies a pointer to it.  They are immutable, so everything shares structure.  Nouns are “persistent” data structures, like Clojure's collections.  If one copies a library, the copy is merely a pointer to the library—the library is a \emph{noun}.  If an imported library can be looked up from a build cache, the builder can copy it into a new app's build subject without duplicating it in memory.

But how does one know whether the cached library is valid?  To achieve global (cross-application) deduplication, one needs a referentially transparent build cache, i.e. a build factored as a full description of an input to the build system.  Since the build system is deterministic, if one sees the same input, one knows that i will produce the same output.

@@

As of \#5745 \citep{Monk2022}, Urbit supplies such a referentially transparent build cache.  What this means in practice is that Urbit can have the memory deduplication benefits of dynamic linking while still using static linking.  Since every filesystem snapshot lives at a unique, immutable, authenticated path within Urbit's scry namespace, reproducible builds are possible on every node, a crucial feature for software supply chain security and reliable app distribution.

@@

Developers have long sought to balance the flexibility and portability of static linking with the better system demands of dynamic linking.  Despite care, the balancing act has led from well-managed mainframe code into the current linker spaghetti situation.  Urbit's \lstinline[style=inlinecode]{++ford} build system elegantly solves for the linking problem by promoting structural sharing of objects (nouns) in memory and by utilizing a referentially transparent build cache across desks.  This balances efficiency in code compilation and building with the reliability of solid-state computing.

\bibliographystyle{plainnat}
\bibliography{linking}
\end{document}

>>>>

\addbibresource{mss.bib}

\newcommand{\authorname}{Ted Blackman \patp{rovnys-ricfer} \\ Philip C. Monk \patp{wicdev-wisryt}}
\newcommand{\authorpatp}{}
\newcommand{\affiliation}{Urbit Foundation \\ Tlon Corporation}

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              Manuscript submitted for review.\\
              % {\it Urbit Systems Technical Journal} I:1 (2024):  1–3. \\
              ~ \\
              Address author correspondence to \patp{rovnys-ricfer}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Static vs. Dynamic Linking}
\fancyfoot[LE,RO]{\thepage}

@@

\begin{abstract}
\noindent
Computing systems that utilize library code must either link to a self-provided version (static linking) or a system-supplied version (dynamic linking).  This can lead to memory duplication or dependency hell.  The Urbit \lstinline[style=inlinecode]{++ford} build system elegantly solves for the linking problem by promoting structural sharing of objects (nouns) in memory and by utilizing a referentially transparent build cache.
\end{abstract}

@@

\noindent
\citeauthor{Locklin2022}'s picturesque exposition highlights the “dependency hell” or “DLL hell” that mires modern software development (cf. \citet{Grimes2003}).  From Urbit's perspective as a solid-state computer, another problem with dynamic linking is that it is not deterministic.  Dynamic linking was something of a pact with the devil, permitting efficient memory usage at the price of legibility.

@@

As another example, the Hoon source file with text \lstinline[style=inlinecode]{'[3 0x4]'} compiles to
\begin{lstlisting}[style=listingblock]
[[%cell [%atom %ud ~] [%atom %ux ~]] [3 0x4]]
\end{lstlisting}
\noindent
This is a vase of a \lstinline[style=inlinecode]{%cell} (pair) of unsigned decimal number (\lstinline[style=inlinecode]{%ud}) and unsigned hexadecimal number (\lstinline[style=inlinecode]{%ux}).  If \lstinline[style=inlinecode]{/foo/hoon} is \lstinline[style=inlinecode]{'3'}, \lstinline[style=inlinecode]{/bar/hoon} is \lstinline[style=inlinecode]{'0x4'}, then importing both of those files changes the build subject to (a vase of) \lstinline[style=inlinecode]{[foo=3 bar=0x4 <stdlib>]}.

@@

In Urbit, everything is a “noun” (a binary tree with arbi\-trarily-sized integers at leaves).  If you “copy” a noun \lstinline[style=inlinecode]{foo}, like \lstinline[style=inlinecode]{[foo foo]}, the runtime just copies a pointer to it.  They are immutable, so everything shares structure.  Nouns are “persistent” data structures, like Clojure's collections.  If one copies a library, the copy is merely a pointer to the library—the library is a \emph{noun}.  If an imported library can be looked up from a build cache, the builder can copy it into a new app's build subject without duplicating it in memory.

But how does one know whether the cached library is valid?  To achieve global (cross-application) deduplication, one needs a referentially transparent build cache, i.e. a build factored as a full description of an input to the build system.  Since the build system is deterministic, if one sees the same input, one knows that it will produce the same output.

@@

As of \citetpr{Monk2022}, Urbit supplies such a referentially transparent build cache.  What this means in practice is that Urbit can have the memory deduplication benefits of dynamic linking while still using static linking.  Since every filesystem snapshot lives at a unique, immutable, authenticated path within Urbit's scry namespace, reproducible builds are possible on every node, a crucial feature for software supply chain security and reliable app distribution.

@@

\sloppy
Developers have long sought to balance the flexibility and portability of static linking with the better system demands of dynamic linking.  Despite care, the balancing act has led from well-managed mainframe code into the current linker spaghetti situation.  Urbit's \lstinline[style=inlinecode]{++ford} build system elegantly solves for the linking problem by promoting structural sharing of objects (nouns) in memory and by utilizing a referentially transparent build cache across desks.  This balances efficiency in code compilation and building with the reliability of solid-state computing.  \tombstone{}

\selectlanguage{USenglish}
\printbibliography
\end{document}
